Higher-Order unification for free!

### 2024-04-12 paper/main.tex
done 4 (can be spellchecked)

\newcommand{\unifcorrect}[3]{
  % \forall \rho #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    \{#3_1, #3_2\} \subseteq \llambda \Rightarrow
      #3_1 #2 #3_2 \mapsto \rho
        \Rightarrow
          \rho #3_1 #1 \rho #3_2  %
}

\newcommand{\unifcomplete}[3]{
  % \forall #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    \{#3_1, #3_2\} \subseteq \llambda \Rightarrow
      % \forall \rho, %
        \rho #3_1 #1 \rho #3_2  %
          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
}
\newcommand{\maybeeta}{\texorpdfstring{\ensuremath{\Diamond\eta}\xspace}{maybeeta}}
\newcommand{\maybebeta}{\texorpdfstring{\ensuremath{\Diamond\beta_0}\xspace}{maybebeta}}
\newcommand{\notllambda}{\texorpdfstring{\ensuremath{\Diamond\llambda}\xspace}{maybellam}}
%\newcommand{\notllambda}{\ensuremath{\Diamond\beta}\xspace}

@@

\subsection{Notational conventions}\todo{move away}

In \cref{sec:eta,sec:invariant1,sec:beta,sec:llam},
we use the following schema to represent the compilation of the list of \Fo{}
problems \foUnifPb into the \Ho problems \hoUnifPb. \mapStore and \linkStore are
respectively the mapping and the link store.
%
\printAlll
  {{{p_1,p_2},{p_3,p_4}}}
  {{{t_1,t_2},{t_3,t_4}}}
  {{{X_1,A_1,{{x}}},{X_2,A_2,{{y}}}}}
  {{{\eta,\Gamma,a,b}}}

We index each sub-problem, sub-mapping, sub-link with its position in the image
starting from $1$ and counting from left to right, top to bottom. For example,
$\hoUnifPb_2$ corresponds to the \Ho problem $t_3 \Ue t_4$.
The compiled version of each $\foUnifPb_i$ is represented by $\hoUnifPb_i$.

Moreover, to indicate the scope of a \Ho variable, we use that scope as
subscript of the considered variable. For example, $X_{xy}$ is the variable $X$ 
having in scope $x$ and $y$.

@@

Even if the compiler, so far, makes no use 
of the higher-order nature of the meta language (all variables have
an empty scope so far) we can already show the code of \Ue{}.

@@

If \wellb{\foUnifPb} then we can prove that \Uo is a good
``first order'' unification algorithm. Later, when the compiler will
enforce it by \cref{prop:w-enforcing} and the proof will be adjusted.

\begin{lemma}
\Cref{prop:correct-ml,prop:complete-ml} hold for the implementation of \Uo above
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting \Ee is as strong as
\Eo. What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed) we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{} terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case after compilation we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

\begin{theorem}[Fidelity in \wellb] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold
\end{theorem}
\begin{proof}[Proof sketch]
Trivial since \elpiIn{progress1} is a no-op and \fstep and \hstep are the same.
Since all terms are $\beta$-normal and \wellb, \Ue is equivalent to \Uo.
\end{proof}

@@

A first problem we encounter when making unification between terms that are
well behaved is the need to treat higher-order variables. 
%
% make test ONLY=7006 TEX=tex
\printAlll
  {{{\lambda x.(f\appsep (X\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{\lambda x.(f\appsep (A\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{X,A,0}}}
  {{}}

In the example above, we can note that the very basilar compilation given in 
the previous section is not able to make the \Ho unification problem succeeds.
The unification of $T_1$ fails while trying to unifying $A \appsep x$ and $x$.
This is due to the fact that $A \appsep x$ (equivalent to \elpiIn{app[uva A [], x]})
is represented as the application of the variable $A$ to the name $x$.
In order to exploit the higher-order unification algorithm of the meta language,
we need to compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$.

@@

In order to address this problem, we add the following rule before rule~\ref{rule:compapp}.

@@

\noindent
Note that compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
is made of bound variables and the hypothetical rule loaded by \elpiIn{comp-lam}
(see below) grants this property. Also note that this rule generates no links.

\noindent
The only detail worth discussing is the fact that the procedure updates a
substitution, rather than just crafting one as presented in
section~\ref{sec:problem-statement}. The reason is that the algorithm folds
over a term, updating a substitution while it traverses it.\noteE[inline]{explain better}

\paragraph{Decompilation} Since no link is created by the compilation of 
\maybebeta terms, no modification should be done
to the \elpiIn{commit-link} predicate.

@@

\begin{lemma}
  \Cref{prop:correct-ml,prop:complete-ml} hold for the implementation of \Uo
  in \cref{sec:founif}
  \end{lemma}
  \begin{proof}[Proof sketch]
  If we look at the \Fo{} terms, the is one more case interesting cases:
  \begin{itemize}
  \item \elpiIn{fapp[fuva X|L] ~\Uo~s}. In this case
   we have $Y_{\vec{x}} \Ue t$ that succeeds with
   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{x}/\vec{y}]\}$ that in turn
\todo{ste x escono da L}   is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{x}/\vec{y}]\}$.
   Thanks to \ref{clause:beta1}
   $(\lambda \vec{y}.s[\vec{x}/\vec{y}])~\vec{x} \Eo s$.
  \end{itemize}
  \end{proof}

\begin{lemma}[\wellb-enforcement]\label{lem:w-enforcement-maybebeta} Even if $\foUnifPb \cap \maybebeta \ne \emptyset$,
  $\hoUnifPb \cup \maybebeta = \emptyset$
\end{lemma}\noteD{Non capisco}
\begin{proof}[Proof sketch]
  problematic terms are mapped to uva by comp, the problematic fapp node is gone.
\end{proof}

\begin{theorem}[Fidelity in \maybebeta] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold
\end{theorem}
\begin{proof}[Proof sketch] thanks to \cref{lem:w-enforcement-maybebeta}
  it is the same as in \cref{sec:simulation}, even if now
  we really need \Ue{} to deal with \llambda, while before a FO unif
  would have done.
\end{proof}
% \subsection{Example}

>>>>

\newcommand{\unifcorrectX}[4]{
  % \forall \rho #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
   #4 \Rightarrow
      #3_1 #2 #3_2 \mapsto \rho
        \Rightarrow
          \rho #3_1 #1 \rho #3_2  %
}
\newcommand{\unifcorrect}[3]{\unifcorrectX{#1}{#2}{#3}{ \{#3_1, #3_2\} \subseteq \llambda}}

\newcommand{\unifcompleteX}[4]{
  % \forall #3_1 #3_2, %
    % \{#3_1, #3_2\} \subseteq \llambda \Rightarrow %
    #4 \Rightarrow
      % \forall \rho, %
        \rho #3_1 #1 \rho #3_2  %
          \Rightarrow \exists \rho', #3_1 #2 #3_2 \mapsto \rho' \land \rho' \subseteq \rho
}
\newcommand{\unifcomplete}[3]{\unifcompleteX{#1}{#2}{#3}{\{#3_1, #3_2\} \subseteq \llambda}}

\newcommand{\maybeeta}{\texorpdfstring{\ensuremath{\Diamond\eta}\xspace}{maybeeta}}
\newcommand{\maybebeta}{\texorpdfstring{\ensuremath{\Diamond\beta}\xspace}{maybebeta}}
\newcommand{\notllambda}{\texorpdfstring{\ensuremath{\Diamond\llambda}\xspace}{maybellam}}
%\newcommand{\notllambda}{\ensuremath{\Diamond\beta}\xspace}

@@

@@

We already have all the pieces to show the code of \Ue{}.

@@

\noindent
So far the compiler is very basic, it does not really
enforce the terms passed to \hstep are in \wellb,
and indeed makes no use of the higher-order capabilities
of the meta language (all generated
variables have an empty scope).
Still, we can prove that \Uo is a good
``first order'' unification algorithm if the input
already happens to be in \wellb.
Later, when the compiler will enforce \cref{prop:w-enforcing}
and the proof will be adjusted to cover for thew cases.

\begin{lemma}[Properties of \Uo]\label{prop:uniffow} The following properties hold for \Uo:
\begin{gather}
  \unifcorrectX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}\\
  \unifcompleteX{\Ee}{\Ue}{t}{\wellb(\{t_1, t_2\})}
\end{gather}
\end{lemma}
\begin{proof}[Proof sketch]
 In this setting \Ee is as strong as \Eo on ground terms.What we have to show is that whenever two different \Fo{}
terms can be made equal by a substitution $\rho$ (plus the \ref{clause:beta1}
and \ref{clause:beta2} if needed) we can produce this $\rho$ by finding
a $\sigma$ via \Ue{} on the corresponding \Ho terms and by decompiling it.
If we look at the syntax of \Fo{} terms the only interesting case is
\elpiIn{fuva X ~\Uo~s}. In this case after compilation we have
$Y \Ue t$ that succeeds with $\sigma = \{ Y \mapsto t\}$ and
$\sigma$ is decompiled to $\rho = \{ Y \mapsto s\}$ by \cref{prop:comprt}.
\end{proof}

\begin{theorem}[Fidelity in \wellb] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb{(\foUnifPb)}$.
\end{theorem}
\begin{proof}[Proof sketch]
Trivial since \elpiIn{progress1} is a no-op and \fstep and \hstep are the same,
and by \cref{prop:uniffow} \Ue is equivalent to \Uo.
\end{proof}

\subsection{Notational conventions}

In the following sections we adopt this notation to talk about the compiler's
output: \foUnifPb is the input set of problems that compiled to
\hoUnifPb with memory mapping \mapStore and links \linkStore. For example:
%
\printAlll
  {{{p_1,p_2},{p_3,p_4}}}
  {{{t_1,t_2},{t_3,t_4}}}
  {{{X_1,A_1,{{x}}},{X_2,A_2,{{y}}}}}
  {{{\eta,\Gamma,a,b}}}

We index each sub-problem, sub-mapping, sub-link with its position
starting from $1$ and counting from left to right, top to bottom.
For example, $\hoUnifPb_2$ corresponds to the \Ho problem $t_3 \Ue t_4$.
%The compiled version of each $\foUnifPb_i$ is represented by $\hoUnifPb_i$.

@@

In order to make \Uo higher-order we need to
take care of terms in \maybebeta.
In the example below, we can see that the basic compilation given in 
the previous section is not able to make the \Ho unification problem succeeds.
%
% make test ONLY=7006 TEX=tex
\printAlll
  {{{\lambda x.(f\appsep (X\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{\lambda x.(f\appsep (A\appsep x)\appsep a),\lambda x.(f\appsep x\appsep a)}}}
  {{{X,A,0}}}
  {{}}
\todo{fix indentation}
The unification problem $\hoUnifPb_1$ fails while trying to unifying
$A \appsep x$ and $x$, equivalent to \elpiIn{app [uva A [], x]}
versus \elpiIn{x}.
In order to exploit the higher-order unification algorithm of the meta language,
we need to compile the \Fo{} term $X\appsep x$ into the \Ho term $A_x$
that is \elpiIn{uva A [x]}.

@@

We add the following rule before rule~\ref{rule:compapp}, where
\elpiIn{pattern-fragment} is a predicate checking is a list of
terms is a list of distinct names.

@@

\noindent
Note that compiling \elpiIn{Ag} cannot create new mappings nor links, since \elpiIn{Ag}
is made of bound variables and the hypothetical rule 
\cref{rule:complamh}
loaded by \elpiIn{comp-lam}
grants this property.

% The only detail worth discussing is the fact that the procedure updates a
% substitution, rather than just crafting one as presented in
% section~\ref{sec:problem-statement}. The reason is that the algorithm folds
% over a term, updating a substitution while it traverses it.\noteE[inline]{explain better}

\paragraph{Decompilation} Since no link is created by the compilation of 
\maybebeta terms, no modification to the \elpiIn{commit-link}
is needed.

@@

\begin{definition}[$\wellb/_\beta$]
$
\wellb/_\beta(X) \Leftrightarrow \forall t \in \subterm{X}, t \not\in (\maybeeta{} ~\cup~ \notllambda{})
$
\end{definition}

\begin{lemma}[Properties of \Uo]\label{prop:uniffowb} The following properties hold for \Uo
  where
  \begin{gather}
    \unifcorrectX{\Ee}{\Ue}{t}{\wellb/_\beta(\{t_1, t_2\})}\\
    \unifcompleteX{\Ee}{\Ue}{t}{\wellb/_\beta(\{t_1, t_2\})}
  \end{gather}
\end{lemma}

  \begin{proof}[Proof sketch]
  If we look at the \Fo{} terms the is one more interesting case,
  namely \elpiIn{fapp[fuva X|W] ~\Uo~s} when \elpiIn{W}
  are distinct names compiled to $\vec{w}$. In this case
  the \Ho problem is $Y_{\vec{w}} \Ue t$ that succeeds with
   $\sigma = \{ Y_{\vec{y}} \mapsto t[\vec{w}/\vec{y}]\}$ that in turn
   is decompiled to $\rho = \{ Y \mapsto \lambda \vec{y}.s[\vec{w}/\vec{y}]\}$.
   Thanks to \ref{clause:beta1}
   $(\lambda \vec{y}.s[\vec{w}/\vec{y}])~\vec{w} \Eo s$.
  \end{proof}

\begin{lemma}[$\wellb/_\beta$-enforcement]\label{lem:w-enforcement-maybebeta}
  Given two terms $s_1$ and $s_2$ in \maybebeta, if $\exists \rho, \rho s_1 \Eo \rho s_2$,
  then 
  $$
    \C{s_i}{t_i}{m_i}{l_i} ~\mathrm{for}~ i \in \{ 1,2 \} \Rightarrow
    t_1 \Ue t_2 \mapsto \sigma
  $$ 
\end{lemma}
\begin{proof}[Proof sketch]
  problematic terms are mapped to uva by comp,
  the problematic fapp node is gone.
  if F x = t then  F = lam x.t[]
  \todo{finish/maybe W enforcement is not needed here}
\end{proof}

\begin{theorem}[Fidelity in $\wellb/_\beta$] \Fullref{prop:simulation} and
\fullref{prop:fidelity} hold if $\;\wellb/_\beta(\foUnifPb)$
\end{theorem}
\begin{proof}[Proof sketch] thanks to \cref{lem:w-enforcement-maybebeta}
  \Ue is as powerful as \Uo in \maybebeta, as well as in \wellb by
  \cref{prop:uniffowb}.
\end{proof}

% \subsection{Example}
